---
title: 'Data 607: Project 2, Part 3'
author: "Aaron Grzasko"
date: "October 9, 2016"
output: 
    html_document:
        theme: simplex
        highlight: haddock
---
  
---  
  
## Assignment Overview

The purpose of this exercise is to gain additional experience manipulating untidy data, using data sets suggested by students in CUNY's Data 607 class.  
  
In Part III, I use a data set that I originally proposed in the week 5 [discussion forum](https://bbhosted.cuny.edu/webapps/discussionboard/do/message?action=list_messages&forum_id=_1397022_1&nav=discussion_board&conf_id=_1377703_1&course_id=_1310339_1&message_id=_20928088_1#msg__20928088_1Id).  
  
The data was originally found on the [USDA website](https://ndb.nal.usda.gov/ndb/search/list).  
  
## Load Libraries  
  
```{r, warning = FALSE, message=FALSE}
# uncomment script below to install necessary packages
# install.packages(c("tidyr","dplyr","stringr"))

library(tidyr)
library(dplyr)
library(stringr)
library(knitr)
```
  
## Data    
  
The data sets for this exercise involve nutritional information for multiple types of milk.  
  
Information for individual milk types was manually selected from the USDA website and exported to separate csv files.  
 
In this exercise, I will join together these separate files into one R object.  
  
Below, I load in messy files for five different milk types.  
  
  
**Skim Milk**
```{r, comment=NA}
skim_link <- "https://raw.githubusercontent.com/spitakiss/Data607_Project2_3/master/skim_cow_milk.csv"
skim_cow_milk <- read.csv(skim_link, header = TRUE, stringsAsFactors = FALSE, skip = 4)
skim_cow_milk <- tbl_df(skim_cow_milk)
head(skim_cow_milk)
```
  
**Rice Milk**  
```{r, comment=NA}
rice_link <- "https://raw.githubusercontent.com/spitakiss/Data607_Project2_3/master/rice_milk.csv"
rice_milk <- read.csv(rice_link, header = TRUE, stringsAsFactors = FALSE, skip = 4)  
rice_milk <- tbl_df(rice_milk)
head(rice_milk)
```  
  
**Almond Milk**  
```{r, comment=NA}
almond_link <- "https://raw.githubusercontent.com/spitakiss/Data607_Project2_3/master/almond_milk.csv"
almond_milk <- read.csv(almond_link, header = TRUE, stringsAsFactors = FALSE, skip = 4)  
almond_milk <- tbl_df(almond_milk)
head(almond_milk)
```
  
**Soy Milk**  
```{r, comment=NA}
soy_link <- "https://raw.githubusercontent.com/spitakiss/Data607_Project2_3/master/soy_milk.csv"
soy_milk <- read.csv(soy_link, header = TRUE, stringsAsFactors = FALSE, skip = 4)
head(soy_link)
```
  
**Goat Milk**  
```{r, comment=NA}
goat_link <- "https://raw.githubusercontent.com/spitakiss/Data607_Project2_3/master/goat_milk.csv"
goat_milk <- read.csv(goat_link, header = TRUE, stringsAsFactors = FALSE, skip = 3)
head(goat_milk)
```

## Tidy Data  
  
**Steps 1 - 7**  
  
1. *Subset columns*.  Select a small subset of necessary columns for this exercise.  
  
2. *Rename columns*.  Column names are long and convoluted.  Here we simplify the names.  
  
3. *Filter Nutrient Values*.  We filter values to pull only those relevant to the analysis.  
  
4. *Remove spaces and commas*.  We remove these elements from elements in the Nutrient column so that we can perform additional manipulations.  
  
5. *Milk Variable*. Assign a new variable that lists the appropriate milk type.
 
6. *Convert rows to column*.  The elements in the nutrient column are each separate variables.  Use the `spread()` function to convert rows to columns.  

7. *Rename Columns*.  The newly created column variables are long and complex.  Here, we simplify.   
  

```{r, comment=NA}
skim_cow_milk <- skim_cow_milk %>%
                # 1
                select(Nutrient, X1.cup...245.0g) %>%
                # 2
                rename(value = X1.cup...245.0g) %>%
                # 3
                filter(Nutrient %in% c("Protein","Energy","Carbohydrate, by difference","Fiber, total dietary"))%>%
                # 4
                mutate(Nutrient = str_replace_all(Nutrient,pattern = " ", replacement = ""),
                       Nutrient = str_replace_all(Nutrient,pattern = ",", replacement = ""),
                        # 5
                        milk_type = "skim_cow_milk"
                ) %>%
                # 6
                spread(Nutrient, value) %>%
                # 7
                rename( carb = Carbohydratebydifference, cal = Energy, fiber = Fibertotaldietary, protein = Protein)    


rice_milk <- rice_milk %>%
                # 1
                select(Nutrient, X8.0.fl.oz..approximate.weight..1.serving....240.0g) %>%
                # 2
                rename(value = X8.0.fl.oz..approximate.weight..1.serving....240.0g) %>%
                # 3
                filter(Nutrient %in% c("Protein","Energy","Carbohydrate, by difference","Fiber, total dietary"))%>%
                # 4
                mutate(Nutrient = str_replace_all(Nutrient,pattern = " ", replacement = ""),
                Nutrient = str_replace_all(Nutrient,pattern = ",", replacement = ""),
                # 5
                milk_type = "rice_milk"
    ) %>%
    # 6
    spread(Nutrient, value) %>%
    # 7
    rename( carb = Carbohydratebydifference, cal = Energy, fiber = Fibertotaldietary, protein = Protein)    


almond_milk <- almond_milk %>%
    # 1
    select(Nutrient, X8.0.fl.oz...240.0g) %>%
    # 2
    rename(value = X8.0.fl.oz...240.0g) %>%
    # 3
    filter(Nutrient %in% c("Protein","Energy","Carbohydrate, by difference","Fiber, total dietary")) %>%
    # 4
    mutate(Nutrient = str_replace_all(Nutrient,pattern = " ", replacement = ""),
           Nutrient = str_replace_all(Nutrient,pattern = ",", replacement = ""),
           # 5
           milk_type = "almond_milk"
    ) %>%
    # 6
    spread(Nutrient, value) %>%
    # 7
    rename( carb = Carbohydratebydifference, cal = Energy, fiber = Fibertotaldietary, protein = Protein)    

soy_milk <- soy_milk %>%
    # 1
    select(Nutrient, X1.cup...243.0g) %>%
    # 2
    rename(value = X1.cup...243.0g) %>%
    # 3
    filter(Nutrient %in% c("Protein","Energy","Carbohydrate, by difference","Fiber, total dietary")) %>%
    # 4
    mutate(Nutrient = str_replace_all(Nutrient,pattern = " ", replacement = ""),
           Nutrient = str_replace_all(Nutrient,pattern = ",", replacement = ""),
           # 5
           milk_type = "soy_milk"
    ) %>%
    # 6
    spread(Nutrient, value) %>%
    # 7
    rename( carb = Carbohydratebydifference, cal = Energy, fiber = Fibertotaldietary, protein = Protein)

goat_milk <- goat_milk %>%
    # 1
    select(Nutrient, X8.0.oz...244.0g) %>%
    # 2
    rename(value = X8.0.oz...244.0g) %>%
    # 3
    filter(Nutrient %in% c("Protein","Energy","Carbohydrate, by difference","Fiber, total dietary")) %>%
    # 4
    mutate(Nutrient = str_replace_all(Nutrient,pattern = " ", replacement = ""),
           Nutrient = str_replace_all(Nutrient,pattern = ",", replacement = ""),
          # 5
           milk_type = "goat_milk"
    ) %>%
    # 6
    spread(Nutrient, value) %>%
    # 7
    rename( carb = Carbohydratebydifference, cal = Energy, fiber = Fibertotaldietary, protein = Protein)    


```
  
**Step 8**  
8. *Master Table*. Finally, we join the milk tables together into one master table.  
  
```{r, comment=NA}
# 8
milk_table <- bind_rows(skim_cow_milk,almond_milk, goat_milk, rice_milk, soy_milk)
milk_table     

```
  
## Analysis  

We already completed the the following assignment instructions:  
  
* Search the USDA website for five different types of milk (examples could include  almond, cow,  rice, goat, hemp), and combine into one master table.    
  
* Manipulate the table format so that each row represents an observation for a specific type of milk.  Nutritional variables should be displayed as separate fields.  
  
Now we will tackle the following items:  
  
* Calculate the ratio of protein to total calories for each type of milk.  
  
* Calculate the average calorie content per cup across all milk types.  
  
* Sort the observations in ascending order by total calories per cup.  
  
```{r, comment = NA}  
milk_table %>%
    mutate(prot_to_cal = round(protein / cal, 4),
           net_carb = carb - fiber
           ) %>%
    arrange(cal)
```

Finally, we we will tackle the last assignment instruction:  
  
* Calculate net carbohydrates for each milk type (i.e. gross carbohydrates - dietary fiber)  
  
```{r, comment=NA}
summarise(milk_table, avg = mean(cal))
```
  


